---
title: "STAT 633A Final Code: Measles Vaccination Rate"
author: "Lyann Luu & Christopher Nguyen"
format:
  html:
    self-contained: true
execute: 
  echo: true
---

# Install Packages

```{r}
library(ggplot2)
library(tidyverse)
library(dplyr)
library(janitor)
library(gtsummary)
```

# Load Data Set

```{r}
# load dataset
measles <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2020/2020-02-25/measles.csv')

# make sure dataset loaded correctly
head(measles)
```

# Raw Data Visuals

```{r}
boxplot(mmr ~ type, data = measles,
        xlab = "School Type",
        ylab = "Vaccination Rate (MMR)",
        main = "Box Plot of Vaccination Rate vs School Type",
        col = "green")
plot(mmr ~ enroll, data = measles,
     xlab = "Enrollment",
     ylab = "Vaccination Rate (MMR)",
     main = "Scatter Plot of Vaccination Rate vs Enrollment",
     col = "purple")
```

```{r}
#figure out that outlier with an extremely high enrollment and low vaccine rate
outlier <- which(measles$enroll > 6000)
outlier

measles[outlier, ]
```

-\> West Valley School Prek-6, Montana \[due to combined grade system\]

# Data Cleaning/Wrangling

```{r}
measles0 <- measles |> select(!(lat:lng)) # remove latitude & longitude columns
measles1 <- data.frame(unique(measles0)) # remove duplicate observations

# remove NA's
measles2 <- measles1 |> filter(!is.na(measles1$state))
measles3 <- measles2 |> filter(!is.na(measles2$type))
measles4 <- measles3 |> filter(!is.na(measles3$enroll))

# remove MMR values of -1
measles5 <- measles4 |> filter(measles4$mmr != -1)

# remove Kindergarten type
measles6 <- measles5 |> filter(measles5$type != "Kindergarten")

# keep only California state
measles_ca <- measles6 |> filter(measles6$state == "California")
```

# Initial Visualization

```{r}
boxplot(mmr ~ type, data = measles_ca,
        xlab = "School Type",
        ylab = "Vaccination Rate (MMR)",
        main = "Box Plot of Vaccination Rate vs School Type",
        col = "green")

plot(mmr ~ enroll, data = measles_ca,
     xlab = "Enrollment",
     ylab = "Vaccination Rate (MMR)",
     main = "Scatter Plot of Vaccination Rate vs Enrollment",
     col = "purple")

summary(measles_ca$enroll)
```

# Modeling

```{r}
# no transformations
par(mfrow = c(1,2))

lm_ca1 <- lm(mmr ~ type + enroll, data = measles_ca)
summary(lm_ca1)
plot(lm_ca1, 1:2, col = "violetred")
```

```{r}
# log transformation on response
par(mfrow = c(1,2))

lm_ca2 <- lm(log(mmr) ~ type + enroll, data = measles_ca)
summary(lm_ca2)
plot(lm_ca2, 1:2, col = "violetred")
```

# \*New Analysis

# Finding the Optimal Polynomial

```{r}
compute_rmse <- function(y, y_pred) {
  n <- length(y)
  sqrt((1 / n) * sum((y - y_pred)^2))
}
```

```{r}
set.seed(2026)
n <- nrow(measles_ca)

train <- sample(1:n, size = 4566) #training data set -> 70% of data
lm1_train <- lm(mmr ~ enroll, data = measles_ca, subset = train)
lm2_train <- lm(mmr ~ poly(enroll, 2), data = measles_ca, subset = train)
lm3_train <- lm(mmr ~ poly(enroll, 3), data = measles_ca, subset = train)
lm4_train <- lm(mmr ~ poly(enroll, 4), data = measles_ca, subset = train)
lm5_train <- lm(mmr ~ poly(enroll, 5), data = measles_ca, subset = train)
lm6_train <- lm(mmr ~ poly(enroll, 6), data = measles_ca, subset = train)

rmse <- rep(0, 6) #initializing variable to store rmse's
y_test <- measles_ca$mmr[-train] #test data set (remaining data not in training data)
new_x <- data.frame(enroll = measles_ca$enroll[-train])

#computing rmse for each order
y1_pred <- predict(lm1_train, newdata = new_x)
rmse[1] <- compute_rmse(y_test, y1_pred)

y2_pred <- predict(lm2_train, newdata = new_x)
rmse[2] <- compute_rmse(y_test, y2_pred)

y3_pred <- predict(lm3_train, newdata = new_x)
rmse[3] <- compute_rmse(y_test, y3_pred)

y4_pred <- predict(lm4_train, newdata = new_x)
rmse[4] <- compute_rmse(y_test, y4_pred)

y5_pred <- predict(lm5_train, newdata = new_x)
rmse[5] <- compute_rmse(y_test, y5_pred)

y6_pred <- predict(lm6_train, newdata = new_x)
rmse[6] <- compute_rmse(y_test, y6_pred)
```

```{r}
rmse
```

```{r}
par(mar=c(4.5,4,1,1)) # adjust margins
plot(c(1:6), rmse, xlab="Degree of polynomial", ylab="RMSE", type='b')
```

Appears that the 2nd polynomial order is the best version to use

```{r}
#linear model with optimal polynomial
lm_ca3 <- lm(mmr ~ type + poly(enroll, 2), data = measles_ca)
summary(lm_ca3)

par(mfrow = c(1,2))
plot(lm_ca3, 1:2, col = "forestgreen")

AIC(lm_ca3)
```

```{r}
# log transformation on response, with poly 2 on enroll (better in terms of AIC)
par(mfrow = c(1,2))
lm_ca4 <- lm(log(mmr) ~ type + poly(enroll, 2), data = measles_ca)
summary(lm_ca4)
plot(lm_ca4, 1:2, col = "forestgreen")

AIC(lm_ca4)
```

Final Linear Model:

$$\widehat{\text{mmr}} = e^{4.547295 + 0.007533(\text{type}_{\text{public}}) - 0.474730(\text{enroll}) - 3.411462(\text{enroll}^2)}$$

# VIF (Variance Inflation Factor)

To see if there is multi-collinearity issues within our predictor variables

A VIF \> 5 = multi-collinearity issues

```{r}
#load the next code chunk for this to work 
pacman::p_load(faraway)
round(vif(lm_ca4), 2) #used on model with no transformations & outliers kept
```

# Anderson-Darling Test for Normality

"Gives more weight to the tails of the distribution" - help files

```{r}
pacman::p_load(nortest)
lm_resid <- data.frame(rstandard(lm_ca4))
ad.test(lm_resid$rstandard.lm_ca4.) #has to be numeric in order to work
```

A = Measures the weighted distance of the data from normality (in our case, it's big and rejects normality

# Modeling Outliers and Leverage

```{r}
#plot for high leverage / outliers in CA data
ind0 <- which(abs(rstandard(lm_ca1)) > 2)
plot(hatvalues(lm_ca1), rstandard(lm_ca1), 
     xlab = "Leverage", 
     ylab = "Standardized Residuals",
     main = "Outlier and Leverage Point Detection Plot")

n <- nrow(measles_ca)
abline(v = 4/n, lty = 2, col = "red") #threshold for high leverage
abline(h = c(-2,2), lty = 2, col = "blue") #threshold for outliers
```

# \*Old Analysis

```{r}
#outliers removal
ind1 <- which(abs(rstandard(lm_ca2)) > 2)
measles_ca2 <- measles_ca[-ind1, ]

# visualizations after outliers removal
boxplot(mmr ~ type, data = measles_ca2,
        xlab = "School Type",
        ylab = "Vaccination Rate (MMR)",
        main = "Box Plot of Vaccination Rate vs School Type",
        col = "green")
plot(mmr ~ enroll, data = measles_ca2,
     xlab = "Enrollment",
     ylab = "Vaccination Rate (MMR)",
     main = "Box Plot of Vaccination Rate vs Enrollment",
     col = "purple")
```

```{r}
# remove bad leverage point
ind2 <- which(measles_ca2$enroll > 600)
measles_ca3 <- measles_ca2[-ind2, ]

# visualizations after leverage point removal
boxplot(mmr ~ type, data = measles_ca3,
        xlab = "School Type",
        ylab = "Vaccination Rate (MMR)",
        main = "Box Plot of Vaccination Rate vs School Type",
        col = "green")
plot(mmr ~ enroll, data = measles_ca3,
     xlab = "Enrollment",
     ylab = "Vaccination Rate (MMR)",
     main = "Box Plot of Vaccination Rate vs Enrollment",
     col = "purple")
```

```{r}
par(mfrow = c(1,2))
# 3rd order polynomial transformation on enroll
lm_ca5 <- lm(log(mmr) ~ type + poly(enroll, 3), data = measles_ca3)
summary(lm_ca5)

# diagnostic models
plot(lm_ca5, 1:2, col = "violetred")

# AIC calculation
AIC(lm_ca5)
```

```{r}
par(mfrow = c(1,2))
# remove outliers
ind3 <- which(abs(rstandard(lm_ca5)) > 2)
measles_ca4 <- measles_ca3[-ind3, ]

# 2nd order polynomial transformation on enroll w/o outliers
lm_ca6 <- lm(log(mmr) ~ type + poly(enroll, 2), data = measles_ca4)
summary(lm_ca6)

# diagnostic models
plot(lm_ca6, 1:2, col = "violetred")

# AIC calculation
AIC(lm_ca6)
```

```{r}
par(mfrow = c(1,2))
# remove outliers (again)
ind4 <- which(abs(rstandard(lm_ca6)) > 2)
measles_ca5 <- measles_ca4[-ind4, ]

# 2nd order polynomial transformation on enroll w/o outliers
lm_ca7 <- lm(log(mmr) ~ type + poly(enroll, 2), data = measles_ca5)
summary(lm_ca7)

# diagnostic models
plot(lm_ca7, 1:2, col = "violetred")

# AIC calculation
AIC(lm_ca7)
```

```{r}
par(mfrow = c(1,2))
# remove outliers
ind5 <- which(abs(rstandard(lm_ca7)) > 2)
measles_ca6 <- measles_ca5[-ind5, ]

# linear model
lm_ca8 <- lm(log(mmr) ~ type + poly(enroll, 3), data = measles_ca6)
summary(lm_ca8)

# diagnostic models
plot(lm_ca8, 1:2, col = "violetred")

# AIC calculation
AIC(lm_ca8)
```

```{r}
#final summary statistics after removal of outliers
boxplot(mmr ~ type, data = measles_ca6,
        xlab = "School Type",
        ylab = "Vaccination Rate (MMR)",
        main = "Box Plot of Vaccination Rate vs School Type",
        col = "green")
plot(mmr ~ enroll, data = measles_ca6,
     xlab = "Enrollment",
     ylab = "Vaccination Rate (MMR)",
     main = "Scatter Plot of Vaccination Rate vs Enrollment",
     col = "purple")
```

# Regression Curve

```{r}
#adding a regression line to final scatter plot above
ggplot(data = measles_ca6, aes(enroll, mmr)) +
  geom_point(size = 0.5, col = "purple") +
  stat_smooth(method = 'lm', formula = y ~ splines::bs(x,3))
```

# Box-Cox Transformation

```{r}
pacman::p_load(MASS)
boxcox(lm_ca1, lambda = seq(0, 25, by = 0.5)) #creating the graph
```

```{r}
pacman::p_load(car)
summary(powerTransform(lm_ca1)) #figuring out lambda_hat
```

# LaTeX for Presentation

$$\text{log(mmr)} = \beta_0 + \beta_1 (\text{type}_\text{Public}) + \beta_2(\text{enroll}) + \beta_3(\text{enroll}^2) + \beta_4(\text{enroll}^3) + \epsilon$$

$$\widehat{\text{log(mmr)}} = 4.5777 + 0.0009 (\text{type}_\text{Public}) + 0.8254(\text{enroll}) - 0.4311(\text{enroll}^2) + 0.1000(\text{enroll}^3)$$

$$\widehat{\text{mmr}} = e^{4.5777 + 0.0009 (\text{type}_\text{Public}) + 0.8254(\text{enroll}) - 0.4311(\text{enroll}^2) + 0.1000(\text{enroll}^3)}$$
